### 2020-09-08 paper/coloringBergeGraphs.tex
Calculating \vartheta

\section{Coloring Perfect Graphs}

A natural problem for perfect graphs is a problem of coloring them. In 1988 Martin Grötschel et al. published an ellipsoid-method-based polynomial algorithm for coloring perfect graphs \cite{Grtschel1993}. We consider it in section \cref{sec:ellipsoidMethod}. However due to its use of the ellipsoid method this algorithm has been usually considered unpractical \cite{coloringSquareFree,Chudnovsky2003, coloringArtemis}.

@@

\subsection{Information theory background}
\label{sec:ellipsoidMethod}

\subsubsection{Shannon Capacity of a graph}

@@

\begin{wrapfigure}{r}{0.35\textwidth}
	\input{tikzpictures/c5.tex}
	\caption{Example noisy channel}%
  \label{fig:c5}
	% \vspace{-0.5cm}
\end{wrapfigure}

@@

\subsubsection{\Lovasz number}

The following introduction of what has become known as \emph{\Lovasz number} is based on excellent lecture notes by \Lovasz \cite{Lovasz95}. \todo{rewrite this so its not blatantly copied.}

@@

On the other hand each term on the left hand side is $5^{-1/4}$, so the left hand side is equal to $|S|5^{-k/2}$, and so $|S| \leq 5^{k/2}$. Thus we get $\alpha(C_5^k) \leq 5 ^{k/2}$ and $\Theta(C_5) = \sqrt{5}$.

This method extends to any graph $G$ in place of $C_5$. All we have to do is find a orthogonal representation that will give us the best bound. So, we can define the \emph{\Lovasz number} of a graph $G$ as \todo{this equation does not really follow from the thought process above, it is a slightly different definition}: 
$$\vartheta(G) = \min\limits_{c,U} \max\limits_{i\in V} \frac{1}{(c^\intercal u_i)^2},$$

@@

Contrary to \Lovasz's first hope \cite{Lovasz1979} $\vartheta(G)$ does not always equal $\Theta(G)$, it is only an upper bound on it. However, these two are equal for some graphs, including all perfect graphs, as is demonstrated in the \Lovasz "sandwich theorem" (\Cref{thm:sandwich}).

@@

\subsection{Computing \boldmath$\vartheta$}

\TODO{TODO}

\begin{theorem}[\Lovasz "sandwich theorem"]
  \label{thm:sandwich}
  For any graph $G$:
  $$ \omega(G) \leq \vartheta(\overline{G}) \leq \chi(G) $$
\end{theorem}

Because in perfect graphs $\omega(G) = \chi(G)$, we get $\omega(G) = \vartheta(G) = \chi(G)$.

\TODO{TODO}

\subsection{Coloring perfect graph using ellipsoid method}
\label{sec:coloringEllipsoid}
The following is based on \citetitle{Laurent2005} by \citeauthor{Laurent2005} \cite{Laurent2005}.

\subsubsection{Maximum cardinality stable set}

@@

\subsubsection{Stable set intersecting all maximum cardinality cliques}
Next, let us show how to find a stable set intersecting all the maximum cardinality cliques of $G$. 
\begin{alg}
  \label{alg:ssIntersectingCliques}
  Input: A perfect graph $G = (V, E)$.

@@

\subsubsection{Minimum coloring}

@@

\subsection{Classical algorithms}
\label{sec:classicalColoring}

>>>>

A natural problem for perfect graphs is a problem of coloring them. In 1988 Martin Grötschel et al. published an ellipsoid-method-based polynomial algorithm for coloring perfect graphs \cite{Grtschel1993}. We consider it in \cref{sec:coloringEllipsoid}. However due to its use of the ellipsoid method this algorithm has been usually considered unpractical \cite{coloringSquareFree,Chudnovsky2003, coloringArtemis}.

@@

\section{Information theory background}
\label{sec:InformationTheory}

\Cref{sec:InformationTheory} and \cref{sec:computingTheta} are based on excellent lecture notes by Lovász \cite{Lovasz95}.

@@

\subsection{Shannon Capacity of a graph}

\begin{wrapfigure}{r}{0.35\textwidth}
  \input{tikzpictures/c5.tex}
  \caption{An example of a noisy channel}%
  \label{fig:c5}
  % \vspace{-0.5cm}
\end{wrapfigure}

@@

\subsection{Lovász number}

@@

On the other hand each term on the left hand side is $5^{-1/4}$, so the left hand side is equal to $|S|5^{-k/2}$, and so $|S| \leq 5^{k/2}$. Since $|S|$ was an arbitrary stable set, we get $\alpha(C_5^k) \leq 5 ^{k/2}$ and $\Theta(C_5) = \sqrt{5}$.

It turns out that this method extends to any graph $G$ in place of $C_5$. All we have to do is find a orthogonal representation that will give us the best bound. So, we can define the \emph{Lovász number} of a graph $G$ as \todo{this equation does not really follow from the thought process above, it is a slightly different definition}:
$$\vartheta(G) = \min\limits_{c,U} \max\limits_{i\in V} \frac{1}{(c^\intercal u_i)^2},$$

@@

Contrary to Lovász's first hope \cite{Lovasz1979} $\vartheta(G)$ does not always equal $\Theta(G)$, it is only an upper bound on it. However, these two are equal for some graphs, including all perfect graphs, as is demonstrated in the Lovász "sandwich theorem".

\begin{theorem}[Lovász "sandwich theorem" \cite{Knuth1994}]
  \label{thm:sandwich}
  For any graph $G$:
  $$ \omega(G) \leq \vartheta(\overline{G}) \leq \chi(G) $$
\end{theorem}

Because in perfect graphs $\omega(G) = \chi(G)$, we get $\omega(G) = \vartheta(\overline{G}) = \chi(G)$.

Therefore, if for any perfect graph $G$, we could calculate $\vartheta(G)$ and $\vartheta(\overline{G})$, we would get $\omega(G)$, $\chi(G)$ and $\alpha(G)$.

@@

\section{Computing \boldmath$\vartheta$}
\label{sec:computingTheta}

First, let us recall some definitions, with \cite{gilbertstrang2020} as a reference for linear algebra. \todo{do we want more lin. algebra definitions?}

\begin{defn}[eigenvector, eigenvalue]
  Let $A$ be an $n \times n$ real matrix. An \emph{eigenvector} of $A$ is a vector $x$ such that $Ax$ is parallel to $x$. In other words, there is a real or complex number $\lambda$, such that $Ax = \lambda x$. This $\lambda$ is called the \emph{eigenvalue} of $A$ belonging to eigenvector $x$.
\end{defn}

If a matrix $A$ is symmetric \footnote{Matrix $A$ is symmetric iff $A = A^\intercal$}, all the eigenvalues are real.

\begin{defn}[positive semidefinite matrix]
  Let $A$ be an $n \times x$ symmetric matrix. $A$ is \emph{positive semidefinite} if all of its eigenvalues are nonnegative. We denote it by $A \succeq 0$.
\end{defn}

We have equivalent definitions of semidefinite matrices.
\begin{theorem}
  For a real symmetric $n \times n$ matrix $A$, the following are equivalent:
  \begin{enumerate}[(i)]
    \item $A$ is positive semidefinite
    \item \label{en:ei2} for every $x \in \mathbb{R}^n$, $x^\intercal Ax$ is nonnegative
    \item for some matrix $U$, $A = U^\intercal U$
    \item $A$ is a nonnegative linear combination of matrices of the type $xx^\intercal$
  \end{enumerate}
\end{theorem}

From (\ref{en:ei2}) it follows that diagonal entries of any positive semidefinite matrix are nonnegative and the sum of two positive semidefinite matrices is positive semidefinite.

We will think of $n \times n$ matrices as vectors with $n^2$ coordinates.
\begin{defn}[convex cone]
  A subset $C$ of $\mathbb{R}^n$ is a \emph{convex cone}, if for any positive scalars $\alpha, \beta$ and for any $x, y \in C$, $\alpha x + \beta y \in C$.
\end{defn}

The fact that the sum of two positive semidefinite matrices is again positive semidefinite, with the fact that every positive scalar multiple of a positive semidefinite matrix is positive semidefinite, translates into the geometric statement that the set of all positive semidefinite matrices forms a convex closed cone $\mathcal{P}_n$ in $\mathbb{R}^{n \times n}$ with vertex 0. This cone $\mathcal{P}_n$ is important but its structure is not trivial.

\paragraph{Semidefinite programs}

Now, we can define a \emph{semidefinite program} to be an optimization problem of the following form:

\begin{equation*}
  \begin{array}{ll@{}ll}
    \text{minimize}   & c^\intercal x                          & \\
    \text{subject to} & x_1A_1 + \ldots + x_nA_n - B \succeq 0   \\
  \end{array}
\end{equation*}
Here $A_1, \ldots, A_n, B$ are given symmetric $m \times m$ matrices and $c \in \mathbb{R}^n$ is a given vector. Any choice of the values $x_i$ that satisfies the given constraint is called a \emph{feasible solution}.

The special case when $A_1, \ldots A_n, B$ are diagonal matrices is a ''generic'' linear program, in fact we can think of semidefinite programs as generalizations of linear programs. Not all properties of linear programs are carried over to semidefinite programs, but the intuition is helpful.

Solving semidefinite programs is a complex topic, we refer to \cite{grotschel1993} for reference. All we need to know is that we can solve semidefinite programs up to an error of $\epsilon$ in polynomial time. One of the methods to do this is called the \emph{ellipsoid method}, hence the name for the coloring algorithm.

\paragraph{Calculating \boldmath$\vartheta$}

Let us recall, that an orthogonal representation of a graph $G = (V, E)$ is a labeling $u: V \rightarrow \mathbb{R}^d$ for some $d$, such that $u_i^\intercal u_j = 0$ for all nonedges $ij$. An \emph{orthonormal} representation is an orthogonal representation with $|u_i| = 1$ for all $i$. The \emph{angle} of an orthogonal representation is the smallest half-angle of a rotational cone containing the representing vectors.

\begin{theorem}[Proposition 5.1 of \cite{Lovasz95}]
  The minimum angle $\phi$ of any orthogonal representation of $G$ is given by $\cos^2\phi = 1/\vartheta(G)$.
\end{theorem}
\TODO{mark all theorems everywhere accordingly}

This leads us to definition of $\vartheta(G)$ in terms of semidefinite programming.

\begin{theorem}[Proposition 5.3 of \cite{Lovasz95}]
  $\vartheta(G)$ is the optimum of the following semidefinite program:

  \begin{equation*}
    \begin{array}{ll@{}ll}
      \text{minimize}   & t      &                                              \\
      \text{subject to} & Y      & \succeq 0                                    \\
                        & Y_{ij} & = -1      & (\forall~ij \in E(\overline{G})) \\
                        & Y_{ii} & = t - 1                                      \\
    \end{array}
  \end{equation*}

  It is also the optimum of the dual program

  \begin{equation*}
    \begin{array}{l@{}rlll}
      \text{maximize~~}   & \sum_{i \in V} \sum_{j \in V}  Z_{ij} &                                     \\
      \text{subject to~~} & Z                                     & \succeq & 0                         \\
                          & Z_{ij}                                & =       & 0 & (\forall~ij \in E(G)) \\
                          & tr(Z)                                 & =       & 1                         \\
    \end{array}
  \end{equation*}
\end{theorem}

Any stable set $S$ of $G$ provides a feasible solution of the sual program, by choosing $Z_{ij} = \frac{1}{S}$, if $i, j \in S$ and 0 otherwise. Similarly, any $k$-coloring of $\overline{G}$ provides a feasible solution of the former semidefinite program, by choosing $Y_{ij} = -1$ if $i$ and $j$ have different colors, $Y_{ii} = k-1$, and $Y_{ij} = 0$ otherwise.

Now, that we know how to calculate $\vartheta(G)$, let us describe the algorithm to calculate the coloring of $G$.

\section{Coloring perfect graph using ellipsoid method}
\label{sec:coloringEllipsoid}
The following is based on \citetitle{Laurent2005} by \citeauthor{Laurent2005} \cite{Laurent2005}.

\subsection{Maximum cardinality stable set}

@@

\subsection{Stable set intersecting all maximum cardinality cliques}
Next, let us show how to find a stable set intersecting all the maximum cardinality cliques of $G$.
\begin{alg}
  \label{alg:ssIntersectingCliques}
  Input: A perfect graph $G = (V, E)$.

@@

\subsection{Minimum coloring}

@@

\section{Classical algorithms}
\label{sec:classicalColoring}

### 2020-09-08 paper/definitions.tex
Calculating \vartheta

\begin{defn}[connected graph, subset]
  A graph $G$ is \emph{connected} if for every pair $\{x, y\} \subseteq V(G)$ of distinct vertices, there is a path from $x$ to $y$. \todo{formal enough?}
  A subset $X \subseteq V(G)$ is connected if the graph $G[X]$ is connected.
\end{defn}

@@

\begin{defn}[Berge graph]
  \label{def:bergeGraph}
  A graph $G$ is \emph{Berge} if both $G$ and $\overline{G}$ have no odd hole as an induced subgraph
\end{defn}

\begin{defn}[perfect graph]
  \label{def:perfectGraph}
  A graph $G$ is \emph{perfect} if for its every induced subgraph $G'$ we have $\chi(G') = \omega(G')$
\end{defn}

>>>>

\begin{defn}[connected graph, subset]
  A graph $G$ is \emph{connected} if for every pair $\{x, y\} \subseteq V(G)$ of distinct vertices, there is a path from $x$ to $y$.
  A subset $X \subseteq V(G)$ is connected if the graph $G[X]$ is connected.
\end{defn}

@@

### 2020-09-08 paper/perfectGraphs.tex
Algos and KT fixes
Calculating \vartheta

Do graphs that admit coloring using only $\omega(G)$ color are "simpler" to further analyze? Not necessarily so. Given a graph $G = (V, E)$, $|V| = n$, let us construct a graph $G'$ equal to the union of $G$ and a clique $K_n$. We can see that indeed $\chi(G') = n = \omega(G')$, but it gives us no indication of the structure of $G$ or $G'$.

A definition of perfect graphs (\cref{def:perfectGraph}) states that given a graph $G$, the chromatic number and cardinality of the largest clique of \emph{its every induced subgraph} should be equal. The notion of perfect graphs was first introduced by Berge in 1961 \cite{CB61} and it indeed captures some of the idea of graph being ''simple'' -- in all perfect graphs the coloring problem, maximum (weighted) clique problem, and maximum (weighted) independent set problem can be solved in polynomial time \cite{grotschel1993}. Other classical NP-complete problems are still NP-complete in perfect graphs e.g. Hamiltonian path \cite{Mller1996}, maximum cut problem \cite{Bodlaender1994} or dominating set problem \cite{Dewdney81}. \todo{all theese citations are about subclasses e.g. biparite graphs. Should we mention some subclasses?}

The most fundamental problem -- the problem of recognizing perfect graphs -- was open since its posing in 1961 until recently. Its solution, a polynomial algorithm recognizing perfect graphs is a union of the strong perfect graph theorem (\cref{sec:SPGT}) stating that a graph is perfect if and only if it is Berge (\cref{def:bergeGraph}) and an algorithm for recognizing Berge graphs in polynomial time (\cref{sec:recognizingBerge}).

@@

The first step to solve the problem of recognizing perfect graphs was the \emph{(weak) perfect graph theorem} first conjured by Berge in 1961~\cite{CB61} and then proven by \Lovasz in 1972~\cite{LL72}.

@@

This theorem is a consequence of a stronger result proven by \Lovasz:
\begin{theorem}
	\label{thm:omegaalpha}
	A graph $G$ is perfect iff for every induced subgraph $H$, the number of vertices of $H$ is at most $\alpha(H)\omega(H)$. \todo{proof?}
\end{theorem}

@@

\begin{theorem}[9.1 of \cite{MC05}]
	\label{thm:91}
	Let $C$ be a shortest odd hole in $G$, with length at least 7. Then there is a relevant triple $(a, b, c)$ of vertices such that
	\begin{itemize}
		\item the set of all $C$-major vertices not in $X(a, b, c)$ is anticonnected
		\item $X(a, b, c) \cap V(C)$ is a subset of the vertex set of some 3-vertex path of $C$.
	\end{itemize}
\end{theorem}
\begin{proof}
	\TODO{do we want it here? 1-2 pages long}
\end{proof}

>>>>

Do graphs that admit coloring using only $\omega(G)$ color are "simpler" to further analyze? Not necessarily so. Given a graph $G = (V, E)$, $|V| = n$, let us construct a graph $G'$ equal to the union of $G$ and a clique $K_n$. We can see that indeed $\chi(G') = n = \omega(G')$, but it gives us no indication of the structure of $G$ or $G'$. This leads us to the definition of \emph{perfect graphs}

\begin{defn}[perfect graph]
  \label{def:perfectGraph}
  A graph $G$ is \emph{perfect} iff for its every induced subgraph $H$ we have $\chi(H) = \omega(H)$
\end{defn}

The notion of perfect graphs was first introduced by Berge in 1961 \cite{CB61} and it indeed captures some of the idea of graph being ''simple'' -- in all perfect graphs the coloring problem, maximum (weighted) clique problem, and maximum (weighted) independent set problem can be solved in polynomial time \cite{grotschel1993}. Other classical NP-complete problems are still NP-complete in perfect graphs e.g. Hamiltonian path \cite{Mller1996}, maximum cut problem \cite{Bodlaender1994} or dominating set problem \cite{Dewdney81}. \todo{all theese citations are about subclasses e.g. biparite graphs. Should we mention some subclasses?}

The most fundamental problem -- the problem of recognizing perfect graphs -- was open since its posing in 1961 until recently. Its solution, a polynomial algorithm recognizing perfect graphs is a union of the strong perfect graph theorem (\cref{sec:SPGT}) stating that a graph is perfect if and only if it is Berge and an algorithm for recognizing Berge graphs in polynomial time (\cref{sec:recognizingBerge}).

\begin{defn}[Berge graph]
  \label{def:bergeGraph}
  A graph $G$ is \emph{Berge} iff both $G$ and $\overline{G}$ have no odd hole as an induced subgraph
\end{defn}

@@

The first step to solve the problem of recognizing perfect graphs was the \emph{(weak) perfect graph theorem} first conjured by Berge in 1961~\cite{CB61} and then proven by Lovász in 1972~\cite{LL72}.

@@

This theorem is a consequence of a stronger result proven by Lovász:
\begin{theorem}
	\label{thm:omegaalpha}
	A graph $G$ is perfect iff for every induced subgraph $H$, the number of vertices of $H$ is at most $\alpha(H)\omega(H)$. \todo{proof?}
\end{theorem}

@@

\begin{theorem}[9.1 of \cite{MC05}]
	\label{thm:91}
	Let $C$ be a shortest odd hole in $G$, with length at least 7. Then there is a relevant triple $(a, b, c)$ of vertices such that
	\begin{itemize}
		\item the set of all $C$-major vertices not in $X(a, b, c)$ is anticonnected
		\item $X(a, b, c) \cap V(C)$ is a subset of the vertex set of some 3-vertex path of $C$.
	\end{itemize}
\end{theorem}
